Bee brain size is positively associated with urban lifestyles but not with positive population trends

### 2022-12-07 Manuscript/Ms.Rmd
phylo signal and starting results

\textbf{Miguel Á. Collado$^{1}$} \orcidlink{0000-0002-4216-317X} \textbf{Jose B. Lanuza$^{2}$} \orcidlink{0000-0002-0287-409X} \textbf{and Ignasi Bartomeus$^{2}$} \orcidlink{0000-0001-7893-4389}

@@

@@

All analyses were undertaken in R version 4.0.5 [@r2021] and the manuscript was written in rmarkdown version 2.17 [@markdown2022].

@@

Brain size was measured as the weight of fixed brain for each specimen. First, all collected individuals were kept in cold and sent them to the laboratory where they were identified by expert taxonomists and anesthetise in cold [@crook2013]. Then, heads were removed with a scalpel and placed in a fixative solution (4% paraformaldehyde in 0.1 M phosphate-buffered saline -PBS- pH = 7.4) to avoid degradation. Subsequently, brains were extracted from the head capsule and placed on a petri dish after removing the retina from the optic lobes and cleaning the entire brain of all tracheae and fat bodies. By doing this, we ensured that our brain measurements mainly accounted for neural tissue. We then placed the brains on a small piece of Parafilm® and dried the exceeding fixative solution with Kimwipes® tissues. Finally, within 4 seconds of the liquid removal, we weighted the brains with the help of a microbalance Sartorius Cubis® and calculated the average weight for the species with more than one specimen.

@@

#Number of species
usa_spp = nrow(records_usa)
eu_spp = nrow(records_eu)
spp_number = nrow(records_eu) + nrow(records_usa)
#Median number of records per dataset
eu_median = median(records_eu$no_rows)
usa_median = median(records_usa$no_rows)

@@

We downloaded occurrence information for all species from the Global Biodiversity Information Facility (GBIF; http://www.gbif.org/) for North America and Europe. The data was downloaded through R programming language with the help of the function *occ_download* from the package *rgbif* version 3.7.3 [@chamberlain2017]. Because the records were biased towards some areas that were associated with political boundaries, we selected the states or countries with highest density of records for our set of species. We selected the boundaries from the maps packages *spData* version 2.2.0 [@spdata] and *giscoR* vesion 0.3.2 [@giscoR], and extracted the occurrences with the help of the function *st_intersection* from the package *sf* version 1.0-7 [@pebesma2018]. For North America, we selected states located on the East Coast of United States (Figure 1A) that covered an approximate area of `r usakm2` km$^{2}$. These states were: Vermont, New Hampshire, Massachusetts, Rhode Island, Connecticut, New York metropolitan area, New Jersey, Delaware and Maryland. Similarly, we selected countries located on the north and center of Europe (Figure 1B) that represented a total area of `r eukm2` km$^{2}$. These countries were: England, Belgium, Netherlands, Germany and Denmark. All occurrence data used in this study can be downloaded from https://doi.org/10.15468/dl.nenuem where just the countries of interest were retrieved. To avoid biases in our analysis, we filtered the occurrence records according the following criteria: (i) we only included species with a widespread distribution of occurrences along our study area; (ii) all species had a minimum number of 50 records; (iii) all duplicated records were excluded (i.e., records of specimens with the same date and location); (iv) we only included georeferenced records with a minimum of two decimals of latitude/longitude coordinates; and (v), we only included records between the time interval 1990 and 2022 in order to avoid large differences with the extracted land cover classifications from 2006. After pruning the data, our final dataset comprised `r spp_number` species with  `r usa_occurrences` occurrence records for United States (`r usa_spp` species; median occurrences = `r usa_median`) and `r eu_occurrences` records for Europe (`r eu_spp` species; median occurrences = `r eu_median`).

@@

Habitat preference was estimated by comparing species occurrences per habitat with the ones that we would expect if they occur at random. For this, we generated from our occurrence matrix 10,000 randomised matrices with help of the function *nullmodel* from the package *bipartite* version 2.16 [@dormann2008]. We implemented this function with the method 'r2dtable' which maintains row and column sums constant by using Patefield's algorithm [@patefield1981]. In other words, it maintains the proportional dominance of the species and habitats constant but reshuffles their associations. We then estimated the percentage of simulated occurrences per species and habitat that were under the observed ones. Finally, we considered that a bee species exhibited statistical preference when the observed occurrences per species and habitat were under the 5th percentile (low preference) or above the 95th percentile (high preference) in comparison with the simulated ones.

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=12, fig.width=12,fig.cap="\\textbf{Figure 2.} Phylogenetic relationship and habitat preferences for the selected bee species with brain weight and intertegular distance information (N=100). A) Phylogenetic tree at the genera level. Tree branches are coloured based on the geographical location of the different bee species (North of Europe, East Coast of United States or from both regions). The deviation of the brain size in relation to the body (i.e., residuals) is indicated with weighted points at the end of the tip branches. Larger points indicate larger brains in proportion to their body size and vice versa. B) Heatmap showing the habitat preference for each bee species. The columns delimit the habitat type (i.e., natural, agricultural, semideveloped and urban) and the rows the different bee species."}

@@

#Plot tree
p <- ggtree(bee.tree100) %<+% brain_weight  +  aes(color=Country)+
    geom_tippoint(aes(size = residuals, color=Country))+
    theme_tree2()+scale_size_continuous(range = c(-1, 3.6),name="Encephalisation")+
    scale_color_manual(name = "Data", values=c("darkgoldenrod1","cyan4","black"),
                       labels = c("Europe", "USA", "USA and Europe"),
                       breaks = c("Europe", "USA", "USA and Europe")) +
  ggtitle("A") + theme(plot.title = element_text(size=18, face="bold"))

@@

p2 = ggplot(d, aes(x=condition, y=Species)) + 
    geom_tile(aes(fill=measurement),color = "white") +
   # scale_fill_gradient(high="black",  low="white") +
    scale_fill_viridis_c(name = "Habitat\n preference") + 
    theme_minimal() +
    xlab(NULL) + 
    ylab(NULL) + 
    theme(axis.text.x = element_text(face="bold", color="black", 
    size=7.5,  vjust=1.15),
    axis.text.y = element_text(face="italic", color="black", size=7.5, hjust=0),
    plot.title = element_text(size=18, face="bold")) +
   ggtitle("B")

@@

all_1 = ggplot(model1, aes(x = residuals, y = estimate__, color=Habitat)) +
geom_point(data =  long_data, aes(x = residuals, y = (Preference)), shape=21, size=2.5) +
geom_line(aes(color=Habitat),size=0.7) +
geom_ribbon(aes(ymin = lower__, ymax = upper__,  fill = Habitat),color=NA, linetype = "dashed",alpha = 0.2)+
theme_bw() +
ylab("Habitat preference") +
xlab("Encephalisation") +
scale_colour_viridis_d() +
scale_fill_viridis_d() +
scale_y_continuous(breaks = seq(0, 1, by = 1),labels = c("Low", "High")) +
ggtitle("A") +
theme(plot.title = element_text(face = "bold", size=18))

@@

#Select just species that fulfilled our criteria of filtering
#Load preferences
preferences_europe = read_csv("../Data/Europe_data/preferences_europe_qualitative.csv") 
preferences_usa = read_csv("../Data/Usa_data/preferences_usa_qualitative.csv")

sp = c(preferences_europe$Species, preferences_usa$Species)

r = r %>% 
mutate(Species = str_replace_all(Species, "_", " ")) %>% 
filter(Species %in% sp)

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=6, fig.width=7, fig.cap="\\textbf{Figure S2.}", fig.pos="p", out.width="80%"}

>>>>

\textbf{Miguel Á. Collado$^{1}$} \orcidlink{0000-0002-4216-317X} \textbf{Jose B. Lanuza$^{2}$} \orcidlink{0000-0002-0287-409X} \textbf{Ferran Sayol$^{3}$} \orcidlink{0000-0003-3540-7487} \textbf{and Ignasi Bartomeus$^{2}$} \orcidlink{0000-0001-7893-4389}

@@

Centre de Recerca Ecològica i Aplicacions Forestals: Bellaterra, Catalonia, Spain

@@

All analyses were undertaken in R version 4.0.5 [@r2021] and the manuscript was written in rmarkdown version 2.17 [@markdown2022]. All data processing and graphics were done with the set of packages from the tidyverse version 1.3.0 [@tidyverse19].

@@

Brain size was measured as the weight of fixed brain for each specimen. First, all collected individuals were kept in cold and sent them to the laboratory where they were identified by expert taxonomists and anesthetise in cold [@crook2013]. Then, heads were removed with a scalpel and placed in a fixative solution (4% paraformaldehyde in 0.1 M phosphate-buffered saline -PBS- pH = 7.4) to avoid degradation. Subsequently, brains were extracted from the head capsule and placed on a petri dish after removing the retina from the optic lobes and cleaning the entire brain of all tracheae and fat bodies. By doing this, we ensured that our brain measurements mainly accounted for neural tissue. We then placed the brains on a small piece of Parafilm® and dried the exceeding fixative solution with Kimwipes® tissues. Then, within 4 seconds of the liquid removal, we weighted the brains with the help of a microbalance Sartorius Cubis®. Finally, we calculated the average weight for the species with more than one specimen. Because weighing bee brains is a highly error-prone task, we excluded values that were under or above 1.5 times the interquartile range for each species.

@@

#Number of species
usa_spp = nrow(records_usa)
eu_spp = nrow(records_eu)
spp_number = nrow(bind_rows(records_eu, records_usa) %>% distinct(species))
spp_shared = nrow(bind_rows(records_eu, records_usa)) - nrow(bind_rows(records_eu, records_usa) %>% distinct(species))

#Median number of records per dataset
eu_median = median(records_eu$no_rows)
usa_median = median(records_usa$no_rows)

@@

We downloaded occurrence information for all species from the Global Biodiversity Information Facility (GBIF; http://www.gbif.org/) for North America and Europe. The data was downloaded through R programming language with the help of the function *occ_download* from the package *rgbif* version 3.7.3 [@chamberlain2017]. Because the records were biased towards some areas that were associated with political boundaries, we selected the states or countries with highest density of records for our set of species. We selected the boundaries from the maps packages *spData* version 2.2.0 [@spdata] and *giscoR* vesion 0.3.2 [@giscoR], and extracted the occurrences with the help of the function *st_intersection* from the package *sf* version 1.0-7 [@pebesma2018]. For North America, we selected states located on the East Coast of United States (Figure 1A) that covered an approximate area of `r usakm2` km$^{2}$. These states were: Vermont, New Hampshire, Massachusetts, Rhode Island, Connecticut, New York metropolitan area, New Jersey, Delaware and Maryland. Similarly, we selected countries located on the north and center of Europe (Figure 1B) that represented a total area of `r eukm2` km$^{2}$. These countries were: England, Belgium, Netherlands, Germany and Denmark. All occurrence data used in this study can be downloaded from https://doi.org/10.15468/dl.5s5kuf where just the countries of interest were retrieved. To avoid biases in our analysis, we filtered the occurrence records according the following criteria: (i) we only included species with a widespread distribution of occurrences along our study area; (ii) all species had a minimum number of 50 records; (iii) all duplicated records were excluded (i.e., records of specimens with the same date and location); (iv) we only included georeferenced records with a minimum of two decimals of latitude/longitude coordinates; and (v), we only included records between the time interval 1990 and 2022 in order to avoid large differences with the extracted land cover classifications from 2006. After pruning the data, our final dataset comprised `r spp_number` species with  `r usa_occurrences` occurrence records for United States (`r usa_spp` species; median occurrences = `r usa_median`) and `r eu_occurrences` records for Europe (`r eu_spp` species; median occurrences = `r eu_median`) wit a total of `r spp_shared` shared species between both regions.

@@

Habitat preference was estimated by comparing species occurrences per habitat with the ones that we would expect if they occur at random. For this, we generated from our occurrence matrix 10,000 randomised matrices with help of the function *nullmodel* from the package *bipartite* version 2.16 [@dormann2008]. We implemented this function with the method 'r2dtable' which maintains row and column sums constant by using Patefield's algorithm [@patefield1981]. In other words, it maintains the proportional dominance of the species and habitats constant but reshuffles their associations. We then estimated the percentage of simulated occurrences per species and habitat that were under the observed ones. Finally, we considered that a bee species exhibited statistical preference when the observed occurrences per species and habitat were under the 20th percentile (low preference) or above the 80th percentile (high preference) in comparison with the simulated ones.

@@

```{r Trait info, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE}
library(dplyr)
#Load brain data
brain_weight = readr::read_csv("../Data/Processing/brain_weight_data.csv")
#Load preferences data
preferences_usa = readr::read_csv("../Data/USA_data/preferences_usa.csv") 
preferences_eu = readr::read_csv("../Data/Europe_data/preferences_europe.csv") 
#Bind rows
pref = bind_rows(preferences_usa, preferences_eu)
#Select unique species
sp = pref %>% distinct(Species) %>% pull()

#Filter species that are in the phylo
brain_weight = brain_weight %>% filter(Species %in% sp)
largest_encephalisation = brain_weight %>% filter(residuals == max(residuals)) %>% select(Species) %>% pull()
lowest_encephalisation = brain_weight %>% filter(residuals == min(residuals)) %>% select(Species) %>% pull()

#Read model output
r2 = read_rds("../Data/Processing/bayessianr2.rds")
r2 = round(r2[1],2)

For the `r nlevels(factor(unique(d$Species)))` bee species, brain size scaled allometrically with body size (Bayesian R$^2$ = `r r2`). However, there were considerable variability in brain size relative to the body across species (Figure S1). Considering this association, the species with the largest and smallest brains relative to the body were the American bee *`r largest_encephalisation`* and the European bee *`r lowest_encephalisation`*, respectively.

Mention how this is constrained by phylogenetic relationship but there is still large variability.

Histogram

Model

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=12, fig.width=12,fig.cap="\\textbf{Figure 2.} Phylogenetic relationship and habitat preferences for the selected bee species with brain weight and intertegular distance information (N=89). A) Phylogenetic tree at the genera level. Tree branches are coloured based on the geographical location of the different bee species (North of Europe, East Coast of United States or from both regions). The deviation of the brain size in relation to the body (i.e., residuals) is indicated with weighted points at the end of the tip branches. Larger points indicate larger brains in proportion to their body size and vice versa. B) Heatmap showing the habitat preference for each bee species. The columns delimit the habitat type (i.e., natural, agricultural, semideveloped and urban) and the rows the different bee species."}

@@

#Plot tree
p <- ggtree(bee.tree100) %<+% brain_weight  +  aes(color=Country)+
    geom_tippoint(aes(size = residuals, color=Country))+
    theme_tree2()+scale_size_continuous(range = c(-1, 4),name="Relative\nbrain size")+
    scale_color_manual(name = "Data", values=c("darkgoldenrod1","cyan4","black"),
                       labels = c("Europe", "USA", "USA and Europe"),
                       breaks = c("Europe", "USA", "USA and Europe")) +
  ggtitle("A") + theme(plot.title = element_text(size=18, face="bold"))

@@

p2 = ggplot(d, aes(x=condition, y=Species)) + 
    geom_tile(aes(fill=measurement),color = "white") +
   # scale_fill_gradient(high="black",  low="white") +
    scale_fill_viridis_c(name = "Habitat\npreference") + 
    theme_minimal() +
    xlab(NULL) + 
    ylab(NULL) + 
    theme(axis.text.x = element_text(face="bold", color="black", 
    size=7.5,  vjust=1.15),
    axis.text.y = element_text(face="italic", color="black", size=7.5, hjust=0),
    plot.title = element_text(size=18, face="bold")) +
   ggtitle("B")

@@

all_1 = ggplot(model1, aes(x = residuals, y = estimate__, color=Habitat)) +
geom_point(data =  long_data, aes(x = residuals, y = (Preference)), shape=21, size=2.5) +
geom_line(aes(color=Habitat),size=0.7) +
geom_ribbon(aes(ymin = lower__, ymax = upper__,  fill = Habitat),color=NA, linetype = "dashed",alpha = 0.2)+
theme_bw() +
ylab("Habitat preference") +
xlab("Relative brain size") +
scale_colour_viridis_d() +
scale_fill_viridis_d() +
scale_y_continuous(breaks = seq(0, 1, by = 1),labels = c("Low", "High")) +
ggtitle("A") +
theme(plot.title = element_text(face = "bold", size=18))

@@

@@

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=6, fig.width=7, fig.cap="\\textbf{Figure S2.}", fig.pos="p", out.width="100%"}
